A general and efficient representation of Ancestral Recombination Graphs

### 2022-04-07 paper.tex

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the precise patterns of
ancestry in intervals along the genome.
We use this framework to define some classifications and metrics of ARG nodes,
and discuss the properties of some recently developed inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

>>>>

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a stochastic process (the coalescent with recombination) together with its
representation, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individual's monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the patterns of
inheritance in intervals along the genome.
We show how this definition can be used to generalise the encoding to
cover both the standard and several simplified versions of ARGs, as well as
enabling efficient access to local trees along the genome. Additionally, we
define some classifications and metrics of ARG nodes. Finally, we relate
these generalisations to the ARGs produced by a number of recently
developed ancestral inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}
